---
title: "Taxonomic predictions"
output: 
  github_document:
    toc: true
    df_print: kable
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(RColorBrewer)
library(MCMCglmm)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```
```{r importdata}
# focusing only on devo
ae_devo <- read.csv(file = "../data/200421_ArrDevFits_filter.csv", header = TRUE)
dat <- read.csv(file = "../data/main_data_rev.csv", header = TRUE)
tree <- read.tree(file = "../data/full_tree_time_calib.nex")
```
```{r}
dat <- select(dat, tpc_num, binomial = parasite_species, genus, fam = family, order, class, phylum, 
              trait_type, num_temps = num_temps_in_curve, cause_of_repeated_measures, same_experiment,
              measurement_metric, simplified_metric,
              sampling_location_available,
              mean_ann_temp, max_month_temp, min_month_temp,
              latitude, altitude_m, author_latitude, dist_zone,
              habitat,
              plant_anim_parasite, endotherm_in_cycle, stage_in_out_host, target_host)

ae_devo <- rename(ae_devo, stderr_arr = SE_BA, stderr_ss = SE_SS)


dat <- left_join(dat, ae_devo, by = 'tpc_num') # join by temperature preference curve identifier

dat <- mutate(dat, E_comb = if_else(best_mod2 == "SS", E_SS, E_Arr),
              se_comb = if_else(best_mod2 == "SS", stderr_ss, stderr_arr))

dat <- filter(dat, !is.na(E_comb)) # remove those without Act E

dat <- mutate(dat, global = if_else(dist_zone == 'global', 'global', 'not global'))

dat <- filter(dat, !(same_experiment == "yes" & simplified_metric != 'mid'), !tpc_num %in% c(61)) # remove non-ind data

dat <- mutate(dat, mean_ann_temp_cen = mean_ann_temp - median(mean_ann_temp),
              max_month_temp_cen = max_month_temp - median(max_month_temp),
              min_month_temp_cen = min_month_temp - median(min_month_temp))

# make habitat variable easy to understand
dat <- mutate(dat, habitat_d1 = if_else(habitat == 'terrestrial', 'Terrestrial',
                                        if_else(habitat == 'freshwater' | habitat == 'marine', 'Aquatic', 'NA')))
dat$habitat_d1[which(dat$habitat_d1 == "NA")] <- NA

```
```{r}
# fix it
tree$tip.label[which(tree$tip.label == "Meloidogyne_hisanica")] <- "Meloidogyne_hispanica"

spp_not_in_dat <- tree$tip.label[!tree$tip.label %in% dat$binomial] # spp in tree excluded from data
# spp_not_in_dat

tree_red <- drop.tip(tree, spp_not_in_dat)

dat$tree_tips <- dat$binomial
```

Number of AE estimates:

```{r}
sum(!is.na(dat$E_comb))
```

Number of species: 

```{r}
n_distinct(dat$binomial)
```

Here's how many groups there are at each level of the taxonomy

```{r}
end_tbl <- dat%>%
  summarise(across(binomial:phylum, n_distinct))%>%
  pivot_longer(everything(), names_to = "tax_level", values_to = "n")
end_tbl
```

Here is AE separated by phylum

```{r}
fx <- ggplot(dat, aes(x = phylum, y = E_comb)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, position = position_jitter(width = 0.33)) +
  labs(x = NULL, y = "Activation energy, eV") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
fx
# ggsave(fx, filename = "../ms_figs/phyla_comp_rev.png", width = 3.5, height = 3.5)
```

Fit series models from the root to the the tip of taxonomy, i.e. first phylum, then add order, then class, etc. Taxonomy is a random effect and there are no fixed effects.

```{r}
# just phyla
prior_tax1 <- list(G=list(G1=list(V=1,nu=0.002),
                          G2=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
# + class
prior_tax2 <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
# + order
prior_tax3 <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,nu=0.002),
                         G4=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
# + fam
prior_tax4 <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,nu=0.002),
                         G4=list(V=1,nu=0.002),
                         G5=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
# + genus
prior_tax5 <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,nu=0.002),
                         G4=list(V=1,nu=0.002),
                         G5=list(V=1,nu=0.002),
                         G6=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
# + species
prior_tax6 <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,nu=0.002),
                         G4=list(V=1,nu=0.002),
                         G5=list(V=1,nu=0.002),
                         G6=list(V=1,nu=0.002),
                         G7=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
```

```{r}
mod1_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax1,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
mod2_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax2,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
mod3_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ order + class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax3,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
mod4_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ fam + order + class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax4,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
mod5_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ genus + fam + order + class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax5,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
mod6_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ binomial + genus + fam + order + class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax6,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=F, verbose = F)
```

From these models, we calculate the variance explained by each new taxonomic level.

```{r}
# the series of models fit
mod_series <- c("phylum", "class", "order", "fam", "genus", "binomial")

# loop through models, calculate R2
i <- 0
if(exists("r2c")){rm(r2c)}
if(exists("r2m")){rm(r2m)}

for(m in list(mod1_tax, mod2_tax, mod3_tax, mod4_tax, mod5_tax, mod6_tax)){

  # calculate fixed effects var
  fixVar <- numeric(length(m$Sol[,1]))
  for(j in 1:length(m$Sol[,1])){
    fixVar_j <- var(as.vector(m$Sol[j,] %*% t(m$X)))
    fixVar[j] <- fixVar_j
  }
  
  # random effects var
  resi <- which(colnames(m$VCV) %in% c('units')) # remove resid variance from variance components
  randVar <- m$VCV[,-resi]
  if(!is.null(colnames(randVar))){ # if only one random effect, it is unnamed and can't use rowSums
    randVar <- rowSums(randVar)-1
    }
  
  # resid var
  resVar <- m$VCV[,'units']
  
  # calculate R2
  r2m <- (fixVar)/(fixVar + randVar + resVar)
  r2c <- (fixVar + randVar)/(fixVar + randVar + resVar)
  cim <- HPDinterval(r2m)
  cic <- HPDinterval(r2c)
  
  if(i == 0){
    r2m_o <- posterior.mode(r2m)
    r2m_l <- cim[1]
    r2m_u <- cim[2]
    r2c_o <- posterior.mode(r2c)
    r2c_l <- cic[1]
    r2c_u <- cic[2]
  } else {
    r2m_o <- c(r2m_o, posterior.mode(r2m))
    r2m_l <- c(r2m_l, cim[1]) 
    r2m_u <- c(r2m_u, cim[2])
    r2c_o <- c(r2c_o, posterior.mode(r2c))
    r2c_l <- c(r2c_l, cic[1]) 
    r2c_u <- c(r2c_u, cic[2])
  }
  i <- i + 1
}
```

And then plot them to see how each taxonomic level explains additional variation in AE

```{r}
end_tbl2 <- data.frame(model = mod_series, r2c_lwr = r2c_l, r2c = r2c_o, r2c_upr = r2c_u)
end_tbl <- left_join(end_tbl, end_tbl2, by = c("tax_level" = "model"))
```
```{r}
end_tbl <- end_tbl%>%
  arrange(r2c)%>%
  mutate(tax_level = if_else(tax_level=="binomial", "species", 
                             if_else(tax_level == "fam", "family", tax_level)))%>%
  mutate(lab = paste0(tax_level, "\n(", as.character(n),")"))
end_tbl <- end_tbl%>%
  mutate(labx = factor(lab, levels = end_tbl$lab),
         group = "group")
```
```{r}
fx <- ggplot(end_tbl, aes(x = labx, y = r2c)) +
  geom_line(aes(group=group), linetype = "dashed") +
  geom_pointrange(aes(x = labx, ymin = r2c_lwr, ymax = r2c_upr), fill = "gray70") +
  labs(y = "Proportion variation explained", x = NULL) +
  scale_y_continuous(limits = c(0,1)) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
        )
fx
```

```{r}
# ggsave(fx, filename = "../ms_figs/Figs_tax_levs.png", units = "in", width = 4, height = 3)
```


It looks like orders and families are most important.

```{r}
# end_tbl%>%
#   mutate(r2_delt = r2c - lag(r2c))
```


```{r}
prior_tax <- list(G=list(G1=list(V=1,nu=0.002),
                         G2=list(V=1,nu=0.002),
                         G3=list(V=1,nu=0.002),
                         G4=list(V=1,nu=0.002),
                         G5=list(V=1,nu=0.002),
                         G6=list(V=1,nu=0.002),
                         G7=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
```

```{r}
mod1_tax <- MCMCglmm(E_comb ~ 1, 
                 random = ~ binomial + genus + fam + order + class + phylum +
                   idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat, 
                 prior = prior_tax,
                 nitt=103000, thin = 50,
                 family = "gaussian", pr=T, verbose = F)
```

```{r}
summary(mod1_tax)
```
Here is the percent variation explained by each taxonomic level in the full model. Family is most important.

```{r}
rv <- rowSums(mod1_tax$VCV[,!colnames(mod1_tax$VCV) %in% c("se_comb.units", "units")])
s0 <- summary(mod1_tax$VCV[,"phylum"]/(rv + mod1_tax$VCV[,"units"]))
s1 <- summary(mod1_tax$VCV[,"class"]/(rv + mod1_tax$VCV[,"units"]))
s2 <- summary(mod1_tax$VCV[,"order"]/(rv + mod1_tax$VCV[,"units"]))
s3 <- summary(mod1_tax$VCV[,"fam"]/(rv + mod1_tax$VCV[,"units"]))
s4 <- summary(mod1_tax$VCV[,"genus"]/(rv + mod1_tax$VCV[,"units"]))

f_tbl <- data.frame(level = c("phylum", "class", "order", "family", "genus"),
                    lwr = c(s0$quantiles[1], s1$quantiles[1], s2$quantiles[1], s3$quantiles[1], s4$quantiles[1]),
                    fit = c(s0$quantiles[3], s1$quantiles[3], s2$quantiles[3], s3$quantiles[3], s4$quantiles[3]),
                    upr = c(s0$quantiles[5], s1$quantiles[5], s2$quantiles[5], s3$quantiles[5], s4$quantiles[5])
                    )
f_tbl
```


```{r}
med_overall <- median(mod1_tax$Sol[,1])
```

Here are predicted values for families (genus and within-species effects marginalized)

```{r}
preds <- predict.MCMCglmm(mod1_tax, marginal = ~ binomial + genus, interval = "confidence")
preds <- data.frame(preds)
names(preds) <- c("ae_fit", "ae.lwr", "ae.upr")

dat_p <- bind_cols(dat, preds)
```
```{r}
fam_p <- dat_p%>%
  group_by(fam)%>%
  summarise(n = n(),
            fit = round(median(ae_fit),2),
            fit.lwr = round(median(ae.lwr),2),
            fit.upr = round(median(ae.upr),2),
            sd_fit = round(sd(ae_fit),3))%>%
  arrange(desc(n), fit)
fam_p
```

```{r}
fam_px <- arrange(fam_p, fit)%>%
  mutate(famx = paste0(fam, " (", as.character(n),")" ))
fam_px <- mutate(fam_px, famx = factor(famx, levels = fam_px$famx))
fam_px <- left_join(
  fam_px, select(dat_p, fam, phylum)%>%distinct()
)

fs2 <- ggplot(fam_px, aes(x = famx, y = fit, color = phylum)) +
  geom_pointrange(aes(ymin = fit.lwr, ymax = fit.upr)) +
  geom_hline(yintercept = med_overall, linetype = "dashed") +
  coord_flip() +
  labs(y = "Activation energy", x = NULL) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(), 
        legend.title = element_blank(),
        legend.background = element_rect(color = 'black'),
        legend.position = c(0.15, 0.85),
        legend.justification = c(0.5,0.5)
        )
fs2
```

Here are predicted values for orders (family, genus and within-species effects marginalized)

```{r}
preds <- predict.MCMCglmm(mod1_tax, marginal = ~ binomial + genus + fam, interval = "confidence")
preds <- data.frame(preds)
names(preds) <- c("ae_fit", "ae.lwr", "ae.upr")

dat_p <- bind_cols(dat, preds)
```
```{r}
ord_p <- dat_p%>%
  group_by(order)%>%
  summarise(n = n(),
            fit = median(ae_fit),
            fit.lwr = median(ae.lwr),
            fit.upr = median(ae.upr),
            sd_fit = round(sd(ae_fit),3))%>%
  arrange(desc(n), fit)
```

```{r}
ord_px <- arrange(ord_p, fit)%>%
  mutate(ordx = paste0(order, " (", as.character(n),")" ))
ord_px <- mutate(ord_px, ordx = factor(ordx, levels = ord_px$ordx))
ord_px <- left_join(
  ord_px, select(dat_p, order, phylum)%>%distinct()
)

fs2b <- ggplot(ord_px, aes(x = ordx, y = fit, color = phylum)) +
  geom_pointrange(aes(ymin = fit.lwr, ymax = fit.upr)) +
  geom_hline(yintercept = med_overall, linetype = "dashed") +
  coord_flip() +
  labs(y = "Activation energy", x = NULL) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(), 
        legend.title = element_blank(),
        legend.background = element_rect(color = 'black'),
        legend.position = c(0.1, 0.85),
        legend.justification = c(0.5,0.5)
        )
fs2b
```



```{r}
library(cowplot)
```

```{r}
fs2x <- fs2 + 
  scale_y_continuous(limits = c(0.24, 1.2)) +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())

fs2y <- fs2b +
  guides(color = F) +
  scale_y_continuous(limits = c(0.24, 1.2)) 
```

```{r}
fs2_comb <- plot_grid(fs2x, fs2y, 
                      ncol = 1, 
                      # rel_heights = c(length(fam_px$fam)/length(ord_px$order), 1),
                      rel_heights = c(1.9, 1))
fs2_comb
```


```{r}
# ggsave(fs2_comb, filename = "../ms_figs/Figs2.png")
```

