---
title: "Testing terms in Table 1"
output: 
  github_document:
    toc: true
    df_print: kable
---

This document fits the series of models analyzing activation energies. The results are summarized in Table 1.

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ape)
library(ggtree)
library(RColorBrewer)
library(MCMCglmm)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```
```{r importdata}
# focusing only on devo
ae_devo <- read.csv(file = "../data/200421_ArrDevFits_filter.csv", header = TRUE)
dat <- read.csv(file = "../data/main_data_rev.csv", header = TRUE)
tree <- read.tree(file = "../data/full_tree_time_calib.nex")
```
```{r}
dat <- select(dat, tpc_num, binomial = parasite_species, genus, fam = family, order, class, phylum, 
              trait_type, num_temps = num_temps_in_curve, cause_of_repeated_measures, same_experiment,
              measurement_metric, simplified_metric,
              sampling_location_available,
              mean_ann_temp, max_month_temp, min_month_temp,
              latitude, altitude_m, author_latitude, dist_zone,
              habitat,
              plant_anim_parasite, endotherm_in_cycle, stage_in_out_host, target_host)

ae_devo <- rename(ae_devo, stderr_arr = SE_BA, stderr_ss = SE_SS)


dat <- left_join(dat, ae_devo, by = 'tpc_num') # join by temperature preference curve identifier

dat <- mutate(dat, E_comb = if_else(best_mod2 == "SS", E_SS, E_Arr),
              se_comb = if_else(best_mod2 == "SS", stderr_ss, stderr_arr))

dat <- filter(dat, !is.na(E_comb)) # remove those without Act E

dat <- mutate(dat, global = if_else(dist_zone == 'global', 'global', 'not global'))

dat <- filter(dat, !(same_experiment == "yes" & simplified_metric != 'mid'), !tpc_num %in% c(61)) # remove non-ind data

dat <- mutate(dat, mean_ann_temp_cen = mean_ann_temp - median(mean_ann_temp),
              max_month_temp_cen = max_month_temp - median(max_month_temp),
              min_month_temp_cen = min_month_temp - median(min_month_temp))

# make habitat variable easy to understand
dat <- mutate(dat, habitat_d1 = if_else(habitat == 'terrestrial', 'Terrestrial',
                                        if_else(habitat == 'freshwater' | habitat == 'marine', 'Aquatic', 'NA')))
dat$habitat_d1[which(dat$habitat_d1 == "NA")] <- NA

```
```{r}
# fix it
tree$tip.label[which(tree$tip.label == "Meloidogyne_hisanica")] <- "Meloidogyne_hispanica"

spp_not_in_dat <- tree$tip.label[!tree$tip.label %in% dat$binomial] # spp in tree excluded from data
# spp_not_in_dat

tree_red <- drop.tip(tree, spp_not_in_dat)

dat$tree_tips <- dat$binomial
```

Number of AE estimates:

```{r}
sum(!is.na(dat$E_comb))
```

Number of species: 

```{r}
n_distinct(dat$binomial)
```


# Terms
## Base model
```{r}
tree_red2 <- makeNodeLabel(tree_red) # standardize node labels
Ainv <- inverseA(tree_red2)$Ainv # make inv of phy cov matrix

dat$tree_tips <- dat$binomial
```
```{r}
prior2<-list(G=list(G1=list(V=1,nu=0.002),
                    G2=list(V=1,nu=0.002),
                    G3=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
iter <- 243000 # number of MCMC iterations used in phylo models throughout
```

```{r}
mod1 <- MCMCglmm(E_Arr ~ 1,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat,
                 prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```

## Development/survival metric

```{r}
mod3 <- MCMCglmm(E_Arr ~ simplified_metric, 
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod3)$solutions
```

Not significant.

# Environmental variables
## Mean temperature

```{r}
mod4 <- MCMCglmm(E_Arr ~ mean_ann_temp_cen, 
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod4)$solutions
```

Not significant.

## Mean temp x exact location 

```{r}
# temp effect on AE depends on author loc and trait type
mod4.3 <- MCMCglmm(E_Arr ~ mean_ann_temp_cen * sampling_location_available, 
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod4.3)$solutions
```

Not significant.

## Temperature variation

```{r}
dat <- mutate(dat, range_temp = max_month_temp - min_month_temp)%>%
  mutate(range_temp_cen = range_temp - median(range_temp))
```
```{r}
mod5 <- MCMCglmm(E_Arr ~ range_temp_cen, 
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod5)$solutions
```

Marg significant.

## Temp variation x exact location

```{r}
mod5xx <- MCMCglmm(E_Arr ~ range_temp_cen * sampling_location_available,                    
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod5xx)$solutions
```

Not significant.

## Latitude

```{r}
dat <- mutate(dat, lat_abs = if_else(latitude == ".", abs(as.integer(author_latitude)), abs(as.integer(latitude))))%>%
  mutate(lat_abs_cen = lat_abs - median(lat_abs))
```
```{r}
mod6 <- MCMCglmm(E_Arr ~ lat_abs_cen,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod6)$solutions
```

Not significant.

## Latitude x global distribution

```{r}
mod7.2 <- MCMCglmm(E_Arr ~ lat_abs_cen*global,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod7.2)$solutions
```

Not significant.

## Habitat - aquatic vs freshwater

```{r}
mod8 <- MCMCglmm(E_Arr ~ habitat_d1,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod8)$solutions
```

Not significant.

## Habitat x latitude 

```{r}
mod8.2 <- MCMCglmm(E_Arr ~ lat_abs_cen*habitat_d1,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod8.2)$solutions
```

Not significant.

# Host-parasite characteristics

## Stage in/out of host

```{r}
mod9 <- MCMCglmm(E_Arr ~ stage_in_out_host,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod9)$solutions
```

Not significant.

## Plant vs animal parasite

```{r}
mod10 <- MCMCglmm(E_Arr ~ plant_anim_parasite,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod10)$solutions
```

Not significant.

## Endotherm in life cycle

```{r}
dat <- dat%>%mutate(endo_ecto2 = if_else(plant_anim_parasite == "plant", "plant", endotherm_in_cycle))
```
```{r}
mod11 <- MCMCglmm(E_Arr ~ endo_ecto2,
                 random = ~ binomial + tree_tips + idh(stderr_arr):units, # w/in spp + phylogenetic effx
                 data = dat, prior = prior2,
                 nitt=iter, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```
```{r}
summary(mod11)$solutions
```

Not significant

# R^2^ Table

```{r}
# the series of models fit
mod_series <- c("base", 
                "metric", "mean temp", "temp x location",
                "temp range", "temp range x location", "latitude",
                "latitude x global", "habitat", "habitat x latitude", 
                "stage in/out of host", "plant vs animal parasite", 
                "endotherm in life cycle?")
```


```{r}
# loop through models, collect DIC and df
i <- 0
if(exists("mod_dic")){rm(mod_dic)}
if(exists("mod_df")){rm(mod_df)}
for(m in list(mod1, 
              mod3, mod4, mod4.3,
              mod5, mod5xx, mod6, 
              mod7.2, mod8, mod8.2,
              mod9, mod10, mod11)){
  if(i == 0){
    mod_dic <- m$DIC
    mod_df <- m$Fixed$nfl + length(m$Random$nrt)-1 # remove 1 df for weights (it is fixed)
  } else {
    mod_dic <- c(mod_dic, m$DIC)
    mod_df <- c(mod_df, m$Fixed$nfl + length(m$Random$nrt)-1)
  }
  i <- i + 1
}

```
```{r}
# loop through models, calculate R2
i <- 0
if(exists("r2c")){rm(r2c)}
if(exists("r2m")){rm(r2m)}

for(m in list(mod1, 
              mod3, mod4, mod4.3,
              mod5, mod5xx, mod6, 
              mod7.2, mod8, mod8.2,
              mod9, mod10, mod11)){

  # calculate fixed effects var
  fixVar <- numeric(length(m$Sol[,1]))
  for(j in 1:length(m$Sol[,1])){
    fixVar_j <- var(as.vector(m$Sol[j,] %*% t(m$X)))
    fixVar[j] <- fixVar_j
  }
  
  # random effects var
  resi <- which(colnames(m$VCV) %in% c('units')) # remove resid variance from variance components
  randVar <- m$VCV[,-resi]
  if(!is.null(colnames(randVar))){ # if only one random effect, it is unnamed and can't use rowSums
    randVar <- rowSums(randVar)-1
    }
  
  # resid var
  resVar <- m$VCV[,'units']
  
  # calculate R2
  r2m <- (fixVar)/(fixVar + randVar + resVar)
  r2c <- (fixVar + randVar)/(fixVar + randVar + resVar)
  cim <- HPDinterval(r2m)
  cic <- HPDinterval(r2c)
  
  if(i == 0){
    r2m_o <- posterior.mode(r2m)
    r2m_l <- cim[1]
    r2m_u <- cim[2]
    r2c_o <- posterior.mode(r2c)
    r2c_l <- cic[1]
    r2c_u <- cic[2]
  } else {
    r2m_o <- c(r2m_o, posterior.mode(r2m))
    r2m_l <- c(r2m_l, cim[1]) 
    r2m_u <- c(r2m_u, cim[2])
    r2c_o <- c(r2c_o, posterior.mode(r2c))
    r2c_l <- c(r2c_l, cic[1]) 
    r2c_u <- c(r2c_u, cic[2])
  }
  i <- i + 1
}
```

```{r}
end_tbl <- data.frame(model = mod_series, df = mod_df)
# end_tbl <- mutate(end_tbl, df_used = df - lag(df, 1))
end_tbl$DIC <- round(mod_dic, 2)
end_tbl$R2m <- paste0(round(r2m_o,2), " [", round(r2m_l,2), "-", round(r2m_u,2), "]")
end_tbl$R2c <- paste0(round(r2c_o,2), " [", round(r2c_l,2), "-", round(r2c_u,2), "]")
```
```{r}
knitr::kable(end_tbl)
```

```{r}

```

