---
title: "Figure 1"
author: "Dan Benesh"
date: "1/9/2020"
output: html_document
---

This notebook creates Fig 1 in the manuscript and it also includes some descriptive stats cited in the first section of the results.

```{r setup, include=FALSE}
library(tidyverse)
library(RColorBrewer)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```
```{r importdata}
# focusing only on devo
ae_devo <- read.csv(file = "../data/200421_ArrDevFits_filter.csv", header = TRUE)
dat <- read.csv(file = "../data/main_data_rev.csv", header = TRUE)
```
```{r}
dat <- select(dat, tpc_num, binomial = parasite_species, genus, fam = family, order, class, phylum, 
              trait_type, num_temps = num_temps_in_curve, cause_of_repeated_measures, same_experiment,
              measurement_metric, simplified_metric,
              sampling_location_available,
              mean_ann_temp, max_month_temp, min_month_temp,
              latitude, altitude_m, author_latitude, dist_zone,
              habitat,
              plant_anim_parasite, endotherm_in_cycle, stage_in_out_host, target_host)
```
```{r}
ae_devo <- rename(ae_devo, stderr_arr = SE_BA, stderr_ss = SE_SS)
```
```{r}
dat <- left_join(dat, ae_devo, by = 'tpc_num') # join by temperature preference curve identifier
dat <- filter(dat, !is.na(E_Arr)) # remove those without Act E
```
```{r}
dat <- mutate(dat, E_comb = if_else(best_mod2 == "SS", E_SS, E_Arr),
              se_comb = if_else(best_mod2 == "SS", stderr_ss, stderr_arr))
```

Number of AE estimates:

```{r}
sum(!is.na(dat$E_comb))
```

Number after removing pseudoreplicates:

```{r}
dat <- filter(dat, !(same_experiment == "yes" & simplified_metric != 'mid'), !tpc_num %in% c(61)) # remove non-ind data
sum(!is.na(dat$E_comb))
```

Number of species:

```{r}
n_distinct(dat$binomial)
```

Number of temps in thermal performance curves

```{r}
summary(dat$num_temps)
```

Median AE estimated by 'best' value

```{r}
m3 <- summarize(dat, 
                q25 = quantile(E_comb, probs = 0.25), 
                med_AE = median(E_comb), 
                q75 = quantile(E_comb, probs = 0.75), 
                n = n(),
                mean_AE = mean(E_comb), 
                skew = e1071::skewness(E_comb),
                kurtosis = e1071::kurtosis(E_comb))
m3
```

Median AE for curves where Arrenhius was best model:

```{r}
filter(dat, best_mod2 == "BA")%>%
   summarize(q25 = quantile(E_comb, probs = 0.25), 
             med_AE = median(E_comb), 
             q75 = quantile(E_comb, probs = 0.75), 
             n = n(),
             mean_AE = mean(E_comb), 
             skew = e1071::skewness(E_comb),
             kurtosis = e1071::kurtosis(E_comb))
```
Median AE for curves where SS was best model:

```{r}
filter(dat, best_mod2 == "SS")%>%
  summarize(q25 = quantile(E_comb, probs = 0.25), 
            med_AE = median(E_comb), 
            q75 = quantile(E_comb, probs = 0.75), 
            n = n(),
            mean_AE = mean(E_comb), 
            skew = e1071::skewness(E_comb),
            kurtosis = e1071::kurtosis(E_comb))
```

Figure 1

```{r}
tg <- ggplot(dat%>%
               mutate(best_mod2 = fct_relevel(best_mod2, c("SS", "BA"))),
                      aes(x = E_comb)) +
  geom_histogram(aes(fill = best_mod2),
                 binwidth = 0.1, color = 'black', 
                 # fill = 'lightgray'
                 ) +
  geom_segment(data = m3, aes(x = med_AE, xend = med_AE, y = 3, yend = 0.3),
               color = 'red', linetype = 'solid', size = 2,
               arrow = arrow(length = unit(0.03, "npc"), type = 'closed')) + # highlight median value
  labs(x = "Activation energy (eV)", y = "Count", fill = NULL) +
  theme(panel.grid.minor = element_blank(),
        legend.background = element_rect(color = "black"),
        legend.position = c(0.95,0.95),
        legend.justification = c(1,1)) +
  scale_fill_manual(values = c("darkgray", "lightgray")) +
  scale_y_continuous(expand = expand_scale(add = c(0,1)))
tg
```
```{r}
tg2 <- ggplot(dat%>%
               mutate(best_mod2 = fct_relevel(best_mod2, c("SS", "BA"))),
                      aes(x = E_comb)) +
  geom_density(aes(E_comb, after_stat(count)),
               bw = 0.1, adjust = 1,
               alpha = 0.1) +
  labs(x = "Activation energy (eV)", y = "Count", fill = NULL) +
  theme(panel.grid.minor = element_blank(),
        legend.background = element_rect(color = "black"),
        legend.position = c(0.95,0.95),
        legend.justification = c(1,1)) +
  scale_fill_manual(values = c("darkgray", "lightgray")) +
  scale_y_continuous(expand = expand_scale(add = c(0,1)))
tg2
```

I like it better together, so let's export a reasonable looking file

```{r}
fig_w <- 4.5
fig_h <- 4
```
```{r}
ggsave(tg, filename = "../ms_figs/Fig1.svg", width = fig_w, height = fig_h, units = 'in')
ggsave(tg2, filename = "../ms_figs/Fig1_density.svg", width = fig_w, height = fig_h, units = 'in')
# ggsave(tg, filename = "../ms_figs/Fig1.svg", width = fig_w, height = fig_h, units = 'in')
```

