---
title: "Figure 3"
output: html_document
---

This notebook creates Fig 3, which combines the env var figs

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(MCMCglmm)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```
```{r importdata}
# focusing only on devo
ae_devo <- read.csv(file = "../data/200421_ArrDevFits_filter.csv", header = TRUE)
dat <- read.csv(file = "../data/main_data_rev.csv", header = TRUE)
tree <- read.tree(file = "../data/full_tree_time_calib.nex")
```
```{r}
dat <- select(dat, tpc_num, binomial = parasite_species, genus, fam = family, order, class, phylum, 
              trait_type, num_temps = num_temps_in_curve, cause_of_repeated_measures, same_experiment,
              measurement_metric, simplified_metric,
              sampling_location_available,
              mean_ann_temp, max_month_temp, min_month_temp,
              latitude, altitude_m, author_latitude, dist_zone,
              habitat,
              plant_anim_parasite, endotherm_in_cycle, stage_in_out_host, target_host)
```
```{r}
ae_devo <- rename(ae_devo, stderr_arr = SE_BA, stderr_ss = SE_SS)
```
```{r}
dat <- left_join(dat, ae_devo, by = 'tpc_num') # join by temperature preference curve identifier
dat <- filter(dat, !is.na(E_Arr)) # remove those without Act E
```
```{r}
dat <- mutate(dat, E_comb = if_else(best_mod2 == "SS", E_SS, E_Arr),
              se_comb = if_else(best_mod2 == "SS", stderr_ss, stderr_arr))
dat <- mutate(dat, global = if_else(dist_zone == 'global', 'global', 'not global'))
```
```{r}
dat <- filter(dat, !(same_experiment == "yes" & simplified_metric != 'mid'), !tpc_num %in% c(61)) # remove non-ind data
```
```{r}
# fix it
tree$tip.label[which(tree$tip.label == "Meloidogyne_hisanica")] <- "Meloidogyne_hispanica"
```

```{r}
spp_not_in_dat <- tree$tip.label[!tree$tip.label %in% dat$binomial] # spp in tree excluded from data
# spp_not_in_dat
```
```{r}
tree_red <- drop.tip(tree, spp_not_in_dat)
```
```{r}
dat$tree_tips <- dat$binomial
```
```{r}
dat <- mutate(dat, mean_ann_temp_cen = mean_ann_temp - median(mean_ann_temp),
              max_month_temp_cen = max_month_temp - median(max_month_temp),
              min_month_temp_cen = min_month_temp - median(min_month_temp))
```
```{r}
# make habitat variable easy to understand
dat <- mutate(dat, habitat_d1 = if_else(habitat == 'terrestrial', 'Terrestrial',
                                        if_else(habitat == 'freshwater' | habitat == 'marine', 'Aquatic', 'NA')))
dat$habitat_d1[which(dat$habitat_d1 == "NA")] <- NA

# sum(is.na(dat$habitat_d))
```
```{r}
dat <- mutate(dat, lat_abs = if_else(latitude == ".", abs(as.integer(author_latitude)), abs(as.integer(latitude))),
              range_temp = max_month_temp - min_month_temp,
              habitat_cen = if_else(habitat_d1 == "Terrestrial", 0.5, -0.5))%>%
  mutate(lat_abs_cen = lat_abs - median(lat_abs), 
         range_temp_cen = range_temp - median(range_temp))
```

First, we'll make a plot for temperature. To put confidence/credible bands on the plot, we need to refit the model and then extract credible intervals. The model includes worm phylogeny, weights, and four environmental covariates (temperature, temperature range, latitude, and habitat).

```{r}
# new data for preds
p <- bind_rows(
  data.frame(binomial = dat$binomial[1], 
             tree_tips = dat$tree_tips[1],
             se_comb = quantile(dat$se_comb, probs = 0.05),
             mean_ann_temp_cen = seq(min(dat$mean_ann_temp_cen)-1,
                                 max(dat$mean_ann_temp_cen)+1,length.out = 50),
             lat_abs_cen = 0,
             range_temp_cen = 0,
             habitat_cen = 0,
             pred = "yes temp"),
  data.frame(binomial = dat$binomial[1], 
             tree_tips = dat$tree_tips[1],
             se_comb = quantile(dat$se_comb, probs = 0.05),
             mean_ann_temp_cen = 0,
             lat_abs_cen = 0, 
             range_temp_cen = seq(min(dat$range_temp_cen)-1,
                                 max(dat$range_temp_cen)+1,length.out = 50),
             habitat_cen = 0,
             pred = "yes temp range"),
  data.frame(binomial = dat$binomial[1], 
             tree_tips = dat$tree_tips[1],
             se_comb = quantile(dat$se_comb, probs = 0.05),
             mean_ann_temp_cen = 0,
             lat_abs_cen = seq(min(dat$lat_abs_cen)-1,
                               max(dat$lat_abs_cen)+1,length.out = 50), # CHANGE!
             range_temp_cen = 0,
             habitat_cen = 0,
             pred = "yes latitude"),
  data.frame(binomial = dat$binomial[1], 
             tree_tips = dat$tree_tips[1],
             se_comb = quantile(dat$se_comb, probs = 0.05),
             mean_ann_temp_cen = 0,
             lat_abs_cen = 0,
             range_temp_cen = 0,
             habitat_cen = c(-0.5, 0.5), habitat_d1 = c("Aquatic", "Terrestrial"),
             pred = "yes habitat")
  )
dat_p <- dat
dat_p$pred <- "no"
dat_p <- bind_rows(dat_p,p)
```

```{r}
tree_red2 <- makeNodeLabel(tree_red) # standardize node labels
Ainv <- inverseA(tree_red2)$Ainv # make inv of phy cov matrix
```
```{r}
prior2<-list(G=list(G1=list(V=1,nu=0.002),
                    G2=list(V=1,nu=0.002),
                    G3=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
```
```{r}
# just mod with temp
mod4 <- MCMCglmm(E_comb ~ mean_ann_temp_cen + lat_abs_cen + range_temp_cen + habitat_cen, 
                 random = ~ binomial + tree_tips + idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat_p, prior = prior2,
                 nitt=243000, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)
```

```{r}
# get predicted values
pdx <- mod4$X # model matrix, just fixed effx
l <- dim(pdx)[1] # number of data points in model
pdx <- pdx[(length(dat$binomial)+1):l,] # restrict to only points where we want preds
num_fe <- mod4$Fixed$nfl # number of fixed effx

# predicted values via matrix multiplication, no taxonomic effx
p_all <- as.matrix(pdx) %*% t(mod4$Sol[,1:num_fe])
# add mean temp to preds, and the make it long
p_all_med <- apply(p_all, MARGIN = 1, median)
p_all_lwr <- apply(p_all, MARGIN = 1, function(x) {quantile(x, probs = 0.025)})
p_all_upr <- apply(p_all, MARGIN = 1, function(x) {quantile(x, probs = 0.975)})

p_all2 <- filter(dat_p, pred != "no")%>%
  mutate(fit = p_all_med,
         upr = p_all_upr,
         lwr = p_all_lwr)
```

Here is the summary of the model used to plot relationships and CIs:

```{r}
summary(mod4)
```


## Temperature

Here is the plot for mean annual temperature.

```{r}
tg <- ggplot(dat, 
             aes(x = mean_ann_temp, y = E_comb)) +
  geom_ribbon(data = filter(p_all2, pred == "yes temp"),
            aes(x = mean_ann_temp_cen + median(dat$mean_ann_temp),
                y = fit, ymin = lwr, ymax = upr),
            alpha = 0.7, fill = 'lightgrey') +
  geom_line(data = filter(p_all2, pred == "yes temp"),
            aes(x = mean_ann_temp_cen + median(dat$mean_ann_temp), 
                y = fit),
            size = 1.5) +
  geom_point(alpha = 0.4, aes(size = 1/se_comb)) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(x = "Mean annual temperature (°C)", y = "Activation energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank())
tg
```
Here is the slope of the line.

```{r}
summary(mod4$Sol[,"mean_ann_temp_cen"])
```
We could also express this as q10, the ratio of rates separated by 10 C. Here is the q10 for 0 and 10 C.

```{r}
r1 <- mod4$Sol[,"(Intercept)"] + mod4$Sol[,"mean_ann_temp_cen"] * (0 - median(dat$mean_ann_temp))
r2 <- mod4$Sol[,"(Intercept)"] + mod4$Sol[,"mean_ann_temp_cen"] * (10 - median(dat$mean_ann_temp))
q10 <- (r2/r1)^(10/(10-0))
summary(q10)
```
Here is the q10 for 10 to 20

```{r}
r1 <- mod4$Sol[,"(Intercept)"] + mod4$Sol[,"mean_ann_temp_cen"] * (10 - median(dat$mean_ann_temp))
r2 <- mod4$Sol[,"(Intercept)"] + mod4$Sol[,"mean_ann_temp_cen"] * (20 - median(dat$mean_ann_temp))
q10 <- (r2/r1)^(10/(20-10))
summary(q10)
```
## Temperature range

Now we'll make the same plot but for temperature range.

```{r}
tg2 <- ggplot(dat, 
             aes(x = range_temp, y = E_comb)) +
  geom_ribbon(data = filter(p_all2, pred == "yes temp range"),
            aes(x = range_temp_cen + median(dat$range_temp),
                y = fit, ymin = lwr, ymax = upr),
            alpha = 0.7, fill = 'lightgrey') +
  geom_line(data = filter(p_all2, pred == "yes temp range"),
            aes(x = range_temp_cen + median(dat$range_temp), 
                y = fit),
            size = 1.5) +
  geom_point(alpha = 0.4, aes(size = 1/se_comb)) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(x = "Range in monthly mean temperatures (°C)", y = "Activation Energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank())
tg2
```
It is very flat. Here is the slope.

```{r}
summary(mod4$Sol[,"range_temp_cen"])
```
## Latitude

Same plot for latitude

```{r}
tg3 <- ggplot(dat, 
             aes(x = lat_abs, y = E_comb)) +
  geom_ribbon(data = filter(p_all2, pred == "yes latitude"),
            aes(x = lat_abs_cen + median(dat$lat_abs),
                y = fit, ymin = lwr, ymax = upr),
            alpha = 0.7, fill = 'lightgrey') +
  geom_line(data = filter(p_all2, pred == "yes latitude"),
            aes(x = lat_abs_cen + median(dat$lat_abs), 
                y = fit),
            size = 1.5) +
  geom_point(alpha = 0.4, aes(size = 1/se_comb)) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(x = "Latitude (° N or S)", y = "Activation energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank())
tg3
```

It is negative. Here is the slope - it was similar to that in the final model. The confidence interval overlaps zero.

```{r}
summary(mod4$Sol[,"lat_abs_cen"])
```
## Habitat

Finally for habitat we plot bars instead of regression lines.

```{r}
tg4 <- ggplot(dat, 
              aes(x = habitat_d1, y = E_comb)) +
  # geom_boxplot() +
  geom_crossbar(data = filter(p_all2, pred == "yes habitat"),
                aes(x = habitat_d1, y = fit, ymin = lwr, ymax = upr),
                alpha = 0.7, fill = 'lightgrey',
                width = 0.3, fatten = 5) +
  geom_point(aes(size = 1/se_comb),
             position = position_jitter(width = 0.3),
             alpha = 0.4) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(x = "Habitat type", y = "Activation energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())  
tg4
```
Here is parameter estimate:

```{r}
summary(mod4$Sol[,"habitat_cen"])
```
```{r}
summary(mod4)
```

Here is the estimated difference:

```{r}
filter(p_all2, pred == "yes habitat")%>%
  select(habitat_cen, habitat_d1, lwr, fit, upr)
```
```{r}
0.84-0.62
```

Here is the difference in the covariates between habitats:

```{r}
dat%>%
  group_by(habitat_d1)%>%
  summarize(mean_ann_temp_cen = median(mean_ann_temp_cen),
            range_temp_cen = median(range_temp_cen),
            lat_abs_cen = median(lat_abs_cen))
```

Since habitats differ in some covariates, we can also check the difference without controlling for other environmental variables.

```{r}
# new data for preds
p <- data.frame(binomial = dat$binomial[1], 
                tree_tips = dat$tree_tips[1],
                se_comb = quantile(dat$se_comb, probs = 0.05),
                habitat_d1 = unique(dat$habitat_d1),
                pred = "yes")

dat_p <- dat
dat_p$pred <- "no"
dat_p <- bind_rows(dat_p,p)
```
```{r}
tree_red2 <- makeNodeLabel(tree_red) # standardize node labels
Ainv <- inverseA(tree_red2)$Ainv # make inv of phy cov matrix
```
```{r}
prior2<-list(G=list(G1=list(V=1,nu=0.002),
                    G2=list(V=1,nu=0.002),
                    G3=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
```
```{r}
# just mod with habitat
mod8 <- MCMCglmm(E_comb ~ habitat_d1, 
                 random = ~ binomial + tree_tips + idh(se_comb):units, # w/in spp + phylogenetic effx
                 data = dat_p, prior = prior2,
                 nitt=23000, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)

```

```{r}
# get predicted values
pdx <- mod8$X # model matrix, just fixed effx
l <- dim(pdx)[1] # number of data points in model
pdx <- pdx[(length(dat$binomial)+1):l,] # restrict to only points where we want preds
num_fe <- mod8$Fixed$nfl # number of fixed effx

# predicted values via matrix multiplication, no taxonomic effx
p_all <- as.matrix(pdx) %*% t(mod8$Sol[,1:num_fe])
# add mean temp to preds, and the make it long
p_all <- bind_cols(select(p, habitat_d1), data.frame(p_all))
p_all <- p_all%>%
  pivot_longer(cols = -habitat_d1,
               names_to = "index", values_to = "pred")
# calculate median fit and upper and lower fits for each value of temp
p_all3 <- group_by(p_all, habitat_d1)%>%
  summarise(fit = median(pred),
            lwr = quantile(pred, probs = 0.025),
            upr = quantile(pred, probs = 0.975))
```

```{r}
tg4justhab <- ggplot(dat, 
             aes(x = habitat_d1, y = E_comb)) +
  # geom_boxplot() +
  geom_crossbar(data = p_all3,
                aes(x = habitat_d1, y = fit, ymin = lwr, ymax = upr),
                alpha = 0.7, fill = 'lightgrey',
                width = 0.3, fatten = 5) +
  geom_point(aes(size = 1/se_comb),
             position = position_jitter(width = 0.3),
             alpha = 0.4) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  labs(x = "Habitat type", y = "Activation energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())  
tg4justhab
```

Here is the estimated difference:

```{r}
p_all3%>%
  select(habitat_d1, lwr, fit, upr)
```

It is pretty similar to when we control for environmental variables. We can also check this result's sensitivity to taxonomy.


```{r}
prior_notax<-list(G=list(G1=list(V=1,fix=1)), # btw observation variance due to measurement errror/SE of curve fit
              R=list(V=1,nu=0.002)) # uninformative prior for within species effect
```
```{r}
# just mod with habitat
mod8_notax <- MCMCglmm(E_comb ~ habitat_d1, 
                 random = ~ idh(se_comb):units, 
                 data = dat_p, prior = prior_notax,
                 nitt=23000, thin = 50,
                 ginverse = list(tree_tips=Ainv),
                 family = "gaussian", pr=F, verbose = F)

```

```{r}
# get predicted values
pdx <- mod8_notax$X # model matrix, just fixed effx
l <- dim(pdx)[1] # number of data points in model
pdx <- pdx[(length(dat$binomial)+1):l,] # restrict to only points where we want preds
num_fe <- mod8_notax$Fixed$nfl # number of fixed effx

# predicted values via matrix multiplication, no taxonomic effx
p_all <- as.matrix(pdx) %*% t(mod8_notax$Sol[,1:num_fe])
# add mean temp to preds, and the make it long
p_all <- bind_cols(select(p, habitat_d1), data.frame(p_all))
p_all <- p_all%>%
  pivot_longer(cols = -habitat_d1,
               names_to = "index", values_to = "pred")
# calculate median fit and upper and lower fits for each value of temp
p_all3 <- group_by(p_all, habitat_d1)%>%
  summarise(fit = median(pred),
            lwr = quantile(pred, probs = 0.025),
            upr = quantile(pred, probs = 0.975))
```

The error is much lower in non-phylogenetic models. It is overconfident.
```{r}
tg4notax <- ggplot(dat, 
             aes(x = habitat_d1, y = E_comb)) +
  # geom_boxplot() +
  geom_crossbar(data = p_all3,
                aes(x = habitat_d1, y = fit, ymin = lwr, ymax = upr),
                alpha = 0.7, fill = 'lightgrey',
                width = 0.3, fatten = 5) +
  geom_point(aes(size = 1/se_comb),
             position = position_jitter(width = 0.3),
             alpha = 0.4) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25)) +
  labs(x = "Habitat type", y = "Activation energy") +
  guides(size = F) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())  
tg4notax
```
The difference is smaller, but the p-value is lower.
```{r}
p_all3%>%
  select(habitat_d1, lwr, fit, upr)
```
```{r}
summary(mod8_notax)
```

## Export
Let's export each plot alone and the combined plots

```{r}
fig_w <- 4.5
fig_h <- 4
```
```{r}
ggsave(tg, filename = "../ms_figs/Fig3w.png", width = fig_w, height = fig_h, units = 'in')
ggsave(tg2, filename = "../ms_figs/Fig3x.png", width = fig_w, height = fig_h, units = 'in')
ggsave(tg3, filename = "../ms_figs/Fig3y.png", width = fig_w, height = fig_h, units = 'in')
ggsave(tg4, filename = "../ms_figs/Fig3z.png", width = fig_w, height = fig_h, units = 'in')
```

Combine plots

```{r}
library(cowplot)
```

```{r}
tg02 <- tg2 + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank())

tg04 <- tg4 + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank())

```

```{r}
f2_c <- plot_grid(tg + labs(y = "Activation energy (eV)"), tg02, 
                  tg3 + labs(y = "Activation energy (eV)"), tg04,
                  align="hv", nrow = 2, 
                  labels = c('(a)', '(b)', '(c)', '(d)')) 
```
```{r}
f2_c
```


```{r}
ggsave(f2_c, filename = "../ms_figs/Fig3_comb.png",
       width = fig_w*2, height = fig_h*2)
ggsave(f2_c, filename = "../ms_figs/Fig3_comb.svg",
       width = fig_w*2, height = fig_h*2)
```


